---
title: 'DATA 605:  Week 7 Assignment'
author: "Aaron Grzasko"
date: "3/18/2017"
output: 
    html_document:
        theme: default
        highlight: haddock
---  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA) 
```

## Problem Set 1

**This week, you’ll have only one programming assignment. Please write a function to compute the expected value and standard deviation of an array of values.  Compare your results with that of R’s mean and std functions.**  
  
First, create a random vector of numbers:    
```{r}
# Random sample of 15 integer values between 1 and 100, with replacement. 
set.seed(1)
sample_data <- sample(100, 15, replace = TRUE)
sample_data
```
  
### Mean Functions  

We can generate the mean (i.e. expected value) of our data in a variety of ways.  
 
Here is a function that calculate the mean of our sample using loops:  

```{r}
loopy_mean <- function(mydata){
  # n = number of data points
  n <- 0
  # sigma = running total sum of data points
  sigma <- 0
  
  # loop through data
  for (val in mydata){
    sigma <- sigma + val
    n <- n + 1
  }
  # mean is equal to sum of values, divided by number of data points
  sigma / n
}

loopy_mean(sample_data)
```

If we want to write a more concise function, we can take advantage of R's built-in `sum()` and `length()` functions:  
```{r}
easier_mean <- function(mydata) return(sum(mydata) / length(mydata))

easier_mean(sample_data)

```

Rather than writing our own function, we can simply take advantage of R's built-in `mean()` function:  
```{r}
mean(sample_data)
```
  
### Standard Deviation Functions  
  
For this homework, we assume we are interested in the sample standard deviation.  
  
Once again, there are multiple approaches to generating this value.  
  
Here is an example using loops:  

```{r}
# function to calcluate std using formula, Sum[(xi - mu)^2] / (n - 1)

loopy_std <- function(mydata){
  # sigma = running sum total of (xi - mu)^2
  sigma <- 0
  
  #n = number of data points
  n <- length(mydata)  
  
  for (val in mydata){
    sigma <- (val - mean(mydata))^2 + sigma
  }
  
  sqrt(sigma / (n-1))
  
}

loopy_std(sample_data)
```
Here is another custom function without loops:     
```{r}
# sample stdev = sqrt([E(X^2) - E(X)^2] * n / (n-1))
easier_std <- function(mydata) {
  sqrt((sum(mydata^2) / length(mydata) - mean(mydata)^2) * length(mydata)/(length(mydata)-1))
}

easier_std(sample_data)
```
Finally, we can calculate standard deviation using the built-in `sd()` function:  
```{r}
sd(sample_data)
```
**Now, consider that instead of being able to neatly fit the values in memory in an array, you have an infinite stream of numbers coming by. How would you estimate the mean and standard deviation of such a stream? Your function should be able to return the current estimate of the mean and standard deviation at any time it is asked. Your program should maintain these current estimates and return them back at any invocation of these functions. (Hint: You can maintain a rolling estimate of the mean and standard deviation and allow these to slowly change over time as you see more and more new values).**
  
The key to creating these rolling estimate functions is to take advantage of R's double arrow assignment operator, "<<-".  
  
Let's first generate some additional sample data to feed to our forthcoming functions:  

```{r}
# create 15 additional random integers between 1 and 100  
set.seed(2)
sample_data2 <- sample(100, 15, replace = TRUE)
sample_data2
```


### Mean Accumulator Function  
  
Here is a function to calculate the rolling mean:
```{r}
# function to return rolling mean
# if no argument is supplied, then return current rolling estimate
# otherwise, update mean with new data

mean_accumulator <- function() {
  # initialize variables for sum of data points (sigma) and number of observations (n)
  # set values to zero if not previously invoked
  sigma <- 0
  n <- 0
  
  # update sigma and n with additional values, if applicable
  function(mydata=NULL){
    if (!is.null(mydata)){
      n <<- n + length(mydata)
      sigma <<- sigma + sum(mydata)
    }
    # re-calculate mean with latest update to sigma and n
    sigma / n
  }
}
```

Now, let's test our function:

```{r}
# create instance of mean_accumulator function
my_mean <- mean_accumulator()  

# calculate mean using first 15 values from sample_data object
my_mean(sample_data)

# update rolling mean with additional data from sample_data2
my_mean(sample_data2)

# calculate mean of combined samples using built-in mean() function
mean(c(sample_data,sample_data2))

# verify that custom accumulator function is identical to mean of combined samples
mean(c(sample_data,sample_data2)) == my_mean()

```
### Standard Deviation Accumulator Function

```{r}
# function to return rolling, sample standard deviation
# if no argument is supplied, then return current rolling estimate
# otherwise, update standard deviation with new data

std_accumulator <- function(){
  
  # initialize variables for sum of data points (sigma), sum of squared data points (sumsq),
  # and number of observations (n)
  # set values to zero if not previously invoked
  
  n <- 0  
  sumsq <- 0
  sigma <- 0
  
  # update sigma, sumsq, and n with additional values, if applicable
  function(mydata=NULL){
    if(!is.null(mydata)){
      sumsq <<- sumsq + sum(mydata^2)
      sigma <<- sigma + sum(mydata) 
      n <<- n + length(mydata)
    }
    
  # return rolling standard deviation
    sqrt((sumsq / n - (sigma / n)^2) * n / (n-1))
  }
}
```

Now, let's test the std_accumulation() function:  
```{r}
# initialize std_accumulator() with new instance
my_std <- std_accumulator()

# calculate stdev using data from sample_data
my_std(sample_data)

# update stdev with data points from sample_data2
my_std(sample_data2)

# calculate stdev using sd() function on combined sample data objects
sd(c(sample_data,sample_data2))

# verify that custom accumulator function is identical to stdev of combined samples   
sd(c(sample_data,sample_data2)) == my_std()
```


### References  
- Hadley Wickham provides an excellent tutorial on functional programming and how to use the double arrow assignment operator in his book, [Advanced R](http://adv-r.had.co.nz/Functional-programming.html).    



